Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-24 paper/section-discussion.tex

We presented a general framework for algebraically indexed types
and instantiated it to yield novel type systems for geometry, taint
tracking, logical information flow and distance-indexed types. Our
framework further demonstrates the power of relational reasoning about
typed programs. From \thmref{thm:abstraction}, we derived
interesting free theorems, type isomorphisms and non-definability
results. We conclude with some observations and suggestions for
further work.

>>>>

We presented a general framework for algebraically indexed types and
instantiated it to yield novel type systems for geometry, logical
information flow and distance-indexed types. Our framework further
demonstrates the power of relational reasoning about typed
programs. From \thmref{thm:abstraction}, we derived interesting free
theorems, type isomorphisms and non-definability results. We conclude
with some observations and suggestions for further work.

### 2012-10-24 paper/section-info-flow.tex

As described by Sabelfeld and Sands \cite{sabelfeld01per}, information
flow can be captured semantically by partial equivalence relations
(PERs). Abadi, Banerjee, Heintze and Riecke \cite{abadi99core} built a
\emph{Core Calculus for Dependency}, using a type system based around
a security level indexed monad $T_lA$, using PERs to prove the
information flow properties. Tse and Zdancewic \cite{tse04translating}
translated Abadi \emph{et al.}'s calculus into System F, translating the
monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
$\alpha_l$, and using Reynolds' abstraction theorem to prove
information flow properties. For example, if the type variable
$\alpha_H$ represents high-level information, then the
non-interference property of the function $f$ could be expressed by
the System F type $A \to (\alpha_H \to B) \to C$. If a program cannot
generate a value of type $\alpha_H$, then it cannot access the value
of $B$, and hence is insensitive to the actual value. Relationships
between security levels are captured by postulating functions
$\alpha_{l_1} \to \alpha_{l_2}$ whenever $l_1$ is a lower security
level than $l_2$.

@@

As a consequence of the relational interpretation given above, it
follows that if we have an index expression $\phi$ that is interpreted
as some value other than $\top$ in a relational environment $\rho$,
then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
\in \rsem{T_\phi\tyPrimNm{bool}}{\rho}$. Thus if a principal is
dissatisfied (i.e.,~$\phi \not= \top$), then data protected by this
principal is indistinguishable from any other data, and a program
cannot get access to the exact value. From this and
\thmref{thm:abstraction}, we obtain the following information flow
result:

\begin{theorem}
  Let $\phi$ and $\psi$ be index expressions of sort $\mathsf{prop}$
  in some indexing context $\Delta$, such that $\psi \not\leq
  \phi$. Then for all programs
  \begin{displaymath}
    \Delta; \Gamma_\infoflow,\Gamma \vdash M : T_\phi\tyPrimNm{bool} \to T_\psi\tyPrimNm{bool}
  \end{displaymath}
  and all terms $N_1,N_2$ of type $T_\phi\tyPrimNm{bool}$, $M\ N_1
  \stackrel{ctx}\approx M\ N_2$.  Thus there is no information flow
  from $M$'s input to its output.
\end{theorem}
Note that if $\psi \leq \phi$, then it is always possible to write the
identity function with this type, using the $\mathrm{up}$ operation.
The theorem also holds if we move to logics other than boolean
logic. For example, if our equational theory models intuitionistic
logic by taking the axioms of Heyting algebras, then the same non-flow
property for programs of type $T_{p\lor\lnot p}\tyPrimNm{bool} \to
T_\top\tyPrimNm{bool}$ holds, due to the lack of excluded middle. If
we take linear logic, then programs of type $T_{p\otimes
  p}\tyPrimNm{bool} \to T_p\tyPrimNm{bool}$ have no information flow
from their input, due to non-provability of $p \vdash p \otimes p$.

>>>>

As described by Sabelfeld and Sands \cite{sabelfeld01per}, information
flow can be captured semantically by partial equivalence relations
(PERs). Abadi, Banerjee, Heintze and Riecke \cite{abadi99core} built a
\emph{Core Calculus for Dependency}, using a type system based around
a security level indexed monad $T_lA$, using PERs to prove the
information flow properties. Tse and Zdancewic \cite{tse04translating}
translated Abadi \emph{et al.}'s calculus into System F\footnote{Tse
  and Zdancewic's translation did not satisfy all the properties that
  they claimed, as pointed out by Shikuma and Igarashi
  \cite{shikuma08proving}. However, this problem is not relevant to
  our discussion here.} translating the monadic type $T_lA$ to
$\alpha_l \to A$ for some free type variable $\alpha_l$, and using
Reynolds' abstraction theorem to prove information flow
properties. For example, if the type variable $\alpha_H$ represents
high-level information, then the non-interference property of the
function $f$ could be expressed by the System F type $A \to (\alpha_H
\to B) \to C$. If a program cannot generate a value of type
$\alpha_H$, then it cannot access the value of $B$, and hence is
insensitive to the actual value. Relationships between security levels
are captured by postulating functions $\alpha_{l_1} \to \alpha_{l_2}$
whenever $l_1$ is a lower security level than $l_2$.

@@

As a consequence of the relational interpretation given above, it
follows that if we have an index expression $\phi$ that is interpreted
as some value other than $\top$ in a relational environment $\rho$,
then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
\in \rsem{T_\phi\tyPrimNm{bool}}{\rho}$. Thus if a principal is
dissatisfied (i.e.,~$\phi \not= \top$), then data protected by this
principal is indistinguishable from any other data, and a program
cannot get access to the exact value. From this observation, and
\thmref{thm:abstraction}, we obtain the following information flow
result:

\begin{theorem}
  Let $\phi$ and $\psi$ be index expressions of sort $\mathsf{prop}$
  in some indexing context $\Delta$, such that $\psi \not\leq
  \phi$. Then for all terms:
  \begin{displaymath}
    \Delta; \Gamma_\infoflow,\Gamma \vdash M : T_\phi\tyPrimNm{bool} \to T_\psi\tyPrimNm{bool}
  \end{displaymath}
  and all terms $N_1,N_2$ of type $T_\phi\tyPrimNm{bool}$, $M\ N_1
  \stackrel{ctx}\approx M\ N_2$.  Thus there is no information flow
  from $M$'s input to its output.
\end{theorem}
Note that if $\psi \leq \phi$, then it is always possible to write the
identity function with this type, using the $\mathrm{up}$ operation.
The theorem also holds if we move to logics other than boolean
logic. For example, if our equational theory models intuitionistic
logic by taking the axioms of Heyting algebras, then the same non-flow
property for programs of type $T_{p\lor\lnot p}\tyPrimNm{bool} \to
T_\top\tyPrimNm{bool}$ holds, due to the lack of excluded middle. If
we take linear logic, then programs of type $T_{p\otimes
  p}\tyPrimNm{bool} \to T_p\tyPrimNm{bool}$ have no information flow
from their input, due to non-provability of $p \vdash p \otimes p$.

### 2012-10-24 paper/section-introduction.tex

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' abstraction theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and abstraction theorem in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}

>>>>

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a
refined type system for information flow based on logic, and a simple
type system with distance-indexed types.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' abstraction theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and most examples in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}

### 2012-10-24 paper/section-metric.tex
Fix minor typo in the distance-indexed types
Revamped section on general non-definability results

\autoref{sec:instantiations} and \autoref{sec:monoid-indexed-types}
presented examples where the relational interpretations of primitive
types relate pairs of elements by some transformation if applying the
transformation to the first element of a pair yields the second. Thus,
the free theorems that we derived directly take the form of
``invariance'' properties, where some equation holds between two
terms. In this section, we examine another instantiation of our
general framework that relates values when they are within a certain
distance. The free theorems that we obtain inform us of the effect
that programs have on the distances between values. For example, a
program $M$ of type:
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$:
\begin{displaymath}
  \begin{array}{l}
    \textrm{if }\abs{x - x'} < \epsilon_1 \textrm{ and } \abs{y - y'} < \epsilon_2 \textrm{ then}\\
    \hspace{2em}\abs{\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'} < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}

\paragraph{Instantiation of the General Framework}
We assume a single indexing sort $\metricSort$ %, intended to 
to represent positive, non-zero real numbers. For the index
operations, we assume the operations $\min, \max, +$ and
multiplication by constant reals. There is a single primitive type
$\tyPrimNm{real}$ with $\primTyArity(\tyPrimNm{real}) =
[\metricSort]$. The primitive operations $\Gamma_{\Metric}$ are as
follows, where $c$ stands for arbitrary real-valued constants:
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
    \underline{c} & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \\
    (+) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    (-) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    \underline{c} * & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to \tyPrim{real}{c\epsilon} \\
    \mathrm{up} & \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\max(\epsilon_1,\epsilon_2)}
  \end{array}
\end{displaymath}
We assume that the index-erasure semantics of the $\tyPrimNm{real}$
type is just the set $\mathbb{R}$, so all except the last operation
have straightforward interpretations. The $\mathrm{up}$ operation is
interpreted just as the identity function. The index-erasure
interpretations of the primitive operations are collected together
into an environment $\eta_\Metric \in \ctxtSem{\Gamma_\Mon}$.

@@

\paragraph{Uniform continuity}
  %By making use of 
Using %the
existential types, % in our framework,
%we can state 
the standard $\epsilon$-$\delta$ definition of
\emph{uniform} continuity can be expressed as %a type:
%\begin{displaymath}
$  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}
For any program $M$ of this type, %we obtain from
\thmref{thm:abstraction} gives a free theorem that is exactly 
%the definition of 
uniform continuity:
\begin{displaymath}
  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. abs{x - x'} < \delta \Rightarrow \abs{\tmSem{M}\ x - \tmSem{M}\ x'} < \epsilon
\end{displaymath}
This definition differs from the $\epsilon$-$\delta$ definition of
(regular) continuity by the order of quantification:
  %For continuity, 
there, $\forall x$ comes before $\exists \delta$, so the distance
$\delta$ may depend on %the value of
$x$. We suspect that to express standard continuity as a type would
require some form of type dependency. Chaudhuri, Gulwani and
Lublinerman \cite{chaudhuri10continuity} have given a program logic
based approach to verifying the continuity of programs.

>>>>

The type system for geometry we discussed in
\autoref{sec:motivating-examples} and \autoref{sec:instantiations}
made use of a relational interpretation of primitive types that
relates pairs of elements by some transformation if applying the
transformation to the first element of a pair yields the second. Thus,
the free theorems that we derived directly take the form of
``invariance'' properties, where some equation holds between two
terms. In this section, we examine another instantiation of our
general framework that relates values when they are within a certain
distance. The free theorems that we obtain inform us of the effect
that programs have on the distances between values. For example, a
program $M$ of type:
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$:
\begin{displaymath}
  \begin{array}{l}
    \textrm{if }\abs{x - x'} < \epsilon_1 \textrm{ and } \abs{y - y'} < \epsilon_2 \textrm{ then}\\
    \hspace{2em}\abs{\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'} < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}

\paragraph{Instantiation of the General Framework}
We assume a single indexing sort $\metricSort$ %, intended to 
to represent positive, non-zero real numbers. For the index
operations, we assume the operations $\min, \max, +$ and
multiplication by constant reals. There is a single primitive type
$\tyPrimNm{real}$ with $\primTyArity(\tyPrimNm{real}) =
[\metricSort]$. The primitive operations $\Gamma_{\Metric}$ are as
follows, where $c$ stands for arbitrary real-valued constants:
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
    \underline{c} & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \\
    (+) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    (-) & \forall \epsilon_1, \epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
    \underline{c} * & \forall \epsilon\mathord:\metricSort.\ \tyPrim{real}{\epsilon} \to \tyPrim{real}{c\epsilon} \\
    \mathrm{up} & \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\max(\epsilon_1,\epsilon_2)}
  \end{array}
\end{displaymath}
We assume that the index-erasure semantics of the $\tyPrimNm{real}$
type is just the set $\mathbb{R}$, so all except the last operation
have straightforward interpretations. The $\mathrm{up}$ operation is
interpreted just as the identity function. The index-erasure
interpretations of the primitive operations are collected together
into an environment $\eta_\Metric \in \ctxtSem{\Gamma_\Metric}$.

@@

\paragraph{Uniform continuity}
  %By making use of 
Using %the
existential types, % in our framework,
%we can state 
the standard $\epsilon$-$\delta$ definition of
\emph{uniform} continuity can be expressed as %a type:
%\begin{displaymath}
$  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}
For any program $M$ of this type, %we obtain from
\thmref{thm:abstraction} gives a free theorem that is exactly 
%the definition of 
uniform continuity:
\begin{displaymath}
  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. \abs{x - x'} \mathord< \delta \Rightarrow \abs{\tmSem{M}\eta_\Metric~x - \tmSem{M}\eta_\Metric~x'} \mathord< \epsilon
\end{displaymath}
This definition differs from the $\epsilon$-$\delta$ definition of
(regular) continuity by the order of quantification:
  %For continuity, 
there, $\forall x$ comes before $\exists \delta$, so the distance
$\delta$ may depend on %the value of
$x$. We suspect that to express standard continuity as a type would
require some form of type dependency. Chaudhuri, Gulwani and
Lublinerman \cite{chaudhuri10continuity} have given a program logic
based approach to verifying the continuity of programs.
